Combining living and fossil taxa into phylogenies: the missing data issue

### 2015-08-24 Manuscript/MPESubmission/Revision/TEM_revised_manuscript2.tex
revisions second round

Here we use a simulation approach to assess the effect of missing data on tree topologies inferred from Total Evidence matrices.
Since the molecular part of a Total Evidence matrix acts like a ``classical" molecular matrix containing only the living taxa \citep{ronquista2012}, the effect of missing data on such matrices is well known \citep{wiensmissing2006,wiensmissing2008,lemmonthe2009,rouresite-specific2011}.
Therefore, we focus only on missing data in the morphological part of the matrix.
We investigate three major parameters that directly affect the completeness and size of the morphological part of the matrix, and reflect empirical biases in data availability: (i) the proportion of living taxa with no morphological data; (ii) the proportion of missing data in the fossil taxa; and (iii) the amount of morphological characters for both living and fossil taxa in the matrix (i.e. the size of the matrix).
We remove data from a Total Evidence matrix by changing the values of these three parameters and then assess how this affects the resulting tree topology.
We infer the topology from the matrices using both Maximum Likelihood and Bayesian inference methods and measure the differences in topology using two different topological distance metrics as proxies for clade conservation and for wild-card taxa placement.
We find that minimizing the number of living taxa with no morphological data and the number of missing morphological characters improves the ability of Total Evidence methods to recover the ``best'' tree topology more so than minimizing the amount of missing data in the fossil record.
Additionally, we find that the ability of Total Evidence methods to recover the ``best'' tree topology is increased when using Bayesian methods.

@@

\subsection{Comparing topologies}
We compared the topology of the ``missing-data'' trees to the ``best'' tree to measure the effect of the three parameters $M_{L}$, $M_{F}$ and $N_{C}$ on tree topology.
We used the Robinson-Foulds distance \citep{RF1981} to assess the number of conserved clade positions and the Triplets distance \citep{dobson1975triplets} to assess the number of wildcard taxa \citep[i.e. taxa that frequently change position in different trees][]{kearneyfragmentary2002}. We used these two metrics because they illustrate two different aspects of tree topology (see Discussion) but also because their performance in measuring differences in topology is well described \citep{Kuhner04112014} and well implemented \citep{Bogdanowicz2012}.
We normalised both metrics using methods described in \citet{Bogdanowicz2012} to generalize our results for any \textit{n} number of taxa.
These metrics are described in detail below.

@@

We measured similarities among the distributions of the different metrics scores (Normalised Robinson-Foulds and Normalised Triplets metric) using the Bhattacharyya Coefficient \citep{Bhattacharyya}.
The Bhattacharyya Coefficient is the probability of overlap between two distributions bounded between 0 (no overlap) and 1 \citep[full overlap;][see Appendix B for calculation details]{Bhattacharyya}.
Note that this is comparable to performing a two-sided t-test, but we use the Bhattacharyya Coefficient here because we are comparing whole distributions not just their means.
When the Bhattacharyya Coefficient between two distributions is $<$0.05, the distributions are significantly different.
When this coefficient is $>$0.95, the distributions are significantly similar.
Values between these two thresholds show the probability of overlap between the distributions but do not allow us to define the significance of the similarity or differences between distributions.
To assess the effect of our missing data parameters, we calculated the Bhattacharyya Coefficient between the distributions of the different metrics scores (Normalised Robinson-Foulds and Normalised Triplets metric) for each pairwise combination of missing data parameters ($M_{L}$, $M_{F}$, $N_{C}$) and parameter states (0\%, 10\%, 25\%, 50\%, 75\% and 100, 90, 75, 50, 25 characters), i.e. $M_{L}$ = 0\%, $M_{F}$ = 0\%, $N_{C}$ = 100; $M_{L}$ = 10\%, $M_{F}$ = 0\%, $N_{C}$ = 100 etc. (see Fig. 2 for more details).
This resulted in 7875 pairwise comparisons (a triangular matrix with $3^5$$\times$$3^5$ cells).
We performed this procedure separately for each tree inference and uncertainty method.
When two combinations of missing data parameters have a similar ability to recover the ``best'' tree topology the Bhattacharyya Coefficient will be close to one.
Conversely, if the two combinations of missing data parameters differ, the Bhattacharyya Coefficient will be close to zero.
Because of the difficulties in representing so many pairwise comparisons in a meaningful way, we summarized these results as a heat map of Bhattacharyya Coefficients (see Fig. 6).
In this type of figure, parameters that have similar effects on recovering the ``best'' topology (either positive or negative effects) will be denoted by similar colour patches in the heat map representation of these comparisons (see Fig. 6).

@@

\subsection{Effects of tree inference methods}
Variation in our ability to recover the ``best'' tree topology depends heavily on the tree inference method (Fig. 4 and Fig. 5).
For morphological data, previous studies have shown some superiority of probabilistic tree inference methods with simple evolutionary models such as the M\textit{kv} model \citep{lewisa2001} over parsimony methods (\citealp{wrightbayesian2014}; but see \citealp{spencerefficacy2013}).
This is, however, the first study, to our knowledge, to compare the performance of the M\textit{kv} model \citep{lewisa2001} for recovering the ``best'' tree topology using Maximum Likelihood and Bayesian methods in a Total Evidence framework.
Our results show that the topology of the Bayesian consensus tree is always closer to the ``best'' tree topology than the ``best'' Maximum Likelihood tree (Fig. 5).
Note that the methodological choice of using the ``true'' tree as a starting tree for the Bayesian Inference rather than a random starting tree (see Methods), had no significant effect on topological recovery (see Appendix A, section ``Effect of the starting tree on Bayesian inference'' for details).
As described above, this is because the Bayesian consensus tree allows a fossil taxon with few data to be placed with a higher confidence at a lower taxonomic level than the Maximum Likelihood tree.
This may also be because the ``best'' Bayesian consensus trees are not completely resolved, thus will always be more similar to the ``missing data" trees than a completely resolved tree like the ``best'' Maximum Likelihood tree.
Nonetheless, we minimized the probability of unresolved ``best'' trees in our Bayesian analyses by only using datasets with strong phylogenetic signal (see Methods).

@@

\bigskip
\noindent
\textbf{Figure 6:} The effects of missing data on topological recovery using Bayesian consensus trees. Both axes show the percentage of missing data from 0\% (white) to 75\% (black) for the three parameters: $M_{L}$ (upper line), $M_{F}$ (middle line) and $N_{C}$ (lower line). The topological recovery is measured as (A) the Normalised Robinson-Foulds metric and (B) the Normalised Triplets metric calculated using the Bhattacharyya Coefficient. The Bhattacharyya Coefficient values are indicated using a color gradient ranging from low probability of overlap in blue, to high probability of overlap in orange. Blue regions denote a poor overlap in Normalised metric values between the different parameter combinations (i.e. the parameters have a strong effect on the metric and thus the topological recovery). Conversely, orange regions denote a high overlap in Normalised metric values between the different parameter combinations (i.e. the parameters have a weak effect on the metric and thus the topological recovery).\\

@@

\noindent
\textbf{Table 1:} Bhattacharyya Coefficients of the pairwise method comparisons.
Each line summarizes the probabilities of overlap between the distributions of the ``best'' tree versus trees from each inference method (Maximum Likelihood; Bayesian consensus; Maximum Likelihood Bootstraps and Bayesian posterior trees) pooled across all combinations of missing data parameter values, using the Normalised Robinson-Foulds (RF) and Triplets (Tr) metrics. 
Values highlighted in bold are the extreme values of high or low probability of overlap between two methods. If two methods have a high probability of overlap, they have a similar ability to recover the ``correct'' tree topology.
Values $>$ 0.95 denote significantly similar distributions and values $<$ 0.05 denote significantly different distributions.\\

>>>>

Here we use a simulation approach to assess the effect of missing data on tree topologies inferred from Total Evidence matrices.
Since the molecular part of a Total Evidence matrix acts like a ``classical" molecular matrix containing only the living taxa \citep{ronquista2012}, the effect of missing data on such matrices is well known \citep{wiensmissing2006,wiensmissing2008,lemmonthe2009,rouresite-specific2011}.
Therefore, we focus only on missing data in the morphological part of the matrix.
We investigate three major parameters that directly affect the completeness and size of the morphological part of the matrix, and reflect empirical biases in data availability: (i) the proportion of living taxa with no morphological data; (ii) the proportion of missing data in the fossil taxa; and (iii) the amount of morphological characters for both living and fossil taxa in the matrix (i.e. the size of the matrix).
We remove data from a Total Evidence matrix by changing the values of these three parameters and then assess how this affects the resulting tree topology.
We infer the topology from the matrices using both Maximum Likelihood and Bayesian inference methods and measure the differences in topology using two different topological distance metrics as proxies for clade conservation and for wild-card taxa placement.
We find that minimising the number of living taxa with no morphological data and the number of missing morphological characters improves the ability of Total Evidence methods to recover the ``best'' tree topology more so than minimising the amount of missing data in the fossil record.
Additionally, we find that the ability of Total Evidence methods to recover the ``best'' tree topology is increased when using Bayesian methods.

@@

\subsection{Comparing topologies}
We compared the topology of the ``missing-data'' trees to the ``best'' tree to measure the effect of the three parameters $M_{L}$, $M_{F}$ and $N_{C}$ on tree topology.
We used the Robinson-Foulds distance \citep{RF1981} to assess the number of conserved clade positions and the Triplets distance \citep{dobson1975triplets} to assess the number of wildcard taxa \citep[i.e. taxa that frequently change position in different trees][]{kearneyfragmentary2002}. We used these two metrics because they illustrate two different aspects of tree topology (see Discussion) but also because their performance in measuring differences in topology is well described \citep{Kuhner04112014} and well implemented \citep{Bogdanowicz2012}.
We normalised both metrics using methods described in \citet{Bogdanowicz2012} to generalise our results for any \textit{n} number of taxa.
These metrics are described in detail below.

@@

We measured similarities among the distributions of the different metrics scores (Normalised Robinson-Foulds and Normalised Triplets metric) using the Bhattacharyya Coefficient \citep{Bhattacharyya}.
The Bhattacharyya Coefficient is the probability of overlap between two distributions bounded between 0 (no overlap) and 1 \citep[full overlap;][see Appendix B for calculation details]{Bhattacharyya}.
Note that this is comparable to performing a two-sided t-test, but we use the Bhattacharyya Coefficient here because we are comparing whole distributions not just their means.
When the Bhattacharyya Coefficient between two distributions is $<$0.05, the distributions are significantly different.
When this coefficient is $>$0.95, the distributions are significantly similar.
Values between these two thresholds show the probability of overlap between the distributions but do not allow us to define the significance of the similarity or differences between distributions.
To assess the effect of our missing data parameters, we calculated the Bhattacharyya Coefficient between the distributions of the different metrics scores (Normalised Robinson-Foulds and Normalised Triplets metric) for each pairwise combination of missing data parameters ($M_{L}$, $M_{F}$, $N_{C}$) and parameter states (0\%, 10\%, 25\%, 50\%, 75\% and 100, 90, 75, 50, 25 characters), i.e. $M_{L}$ = 0\%, $M_{F}$ = 0\%, $N_{C}$ = 100; $M_{L}$ = 10\%, $M_{F}$ = 0\%, $N_{C}$ = 100 etc. (see Fig. 2 for more details).
This resulted in 7875 pairwise comparisons (a triangular matrix with $3^5$$\times$$3^5$ cells).
We performed this procedure separately for each tree inference and uncertainty method.
When two combinations of missing data parameters have a similar ability to recover the ``best'' tree topology the Bhattacharyya Coefficient will be close to one.
Conversely, if the two combinations of missing data parameters differ, the Bhattacharyya Coefficient will be close to zero.
Because of the difficulties in representing so many pairwise comparisons in a meaningful way, we summarised these results as a heat map of Bhattacharyya Coefficients (see Fig. 6).
In this type of figure, parameters that have similar effects on recovering the ``best'' topology (either positive or negative effects) will be denoted by similar colour patches in the heat map representation of these comparisons (see Fig. 6).

@@

\subsection{Effects of tree inference methods}
Variation in our ability to recover the ``best'' tree topology depends heavily on the tree inference method (Fig. 4 and Fig. 5).
For morphological data, previous studies have shown some superiority of probabilistic tree inference methods with simple evolutionary models such as the M\textit{kv} model \citep{lewisa2001} over parsimony methods (\citealp{wrightbayesian2014}; but see \citealp{spencerefficacy2013}).
This is, however, the first study, to our knowledge, to compare the performance of the M\textit{kv} model \citep{lewisa2001} for recovering the ``best'' tree topology using Maximum Likelihood and Bayesian methods in a Total Evidence framework.
Our results show that the topology of the Bayesian consensus tree is always closer to the ``best'' tree topology than the ``best'' Maximum Likelihood tree (Fig. 5).
Note that the methodological choice of using the ``true'' tree as a starting tree for the Bayesian Inference rather than a random starting tree (see Methods), had no significant effect on topological recovery (see Appendix A, section ``Effect of the starting tree on Bayesian inference'' for details).
As described above, this is because the Bayesian consensus tree allows a fossil taxon with few data to be placed with a higher confidence at a lower taxonomic level than the Maximum Likelihood tree.
This may also be because the ``best'' Bayesian consensus trees are not completely resolved, thus will always be more similar to the ``missing data" trees than a completely resolved tree like the ``best'' Maximum Likelihood tree.
Nonetheless, we minimised the probability of unresolved ``best'' trees in our Bayesian analyses by only using datasets with strong phylogenetic signal (see Methods).

@@

\bigskip
\noindent
\textbf{Figure 6:} The effects of missing data on topological recovery using Bayesian consensus trees. Both axes show the percentage of missing data from 0\% (white) to 75\% (black) for the three parameters: $M_{L}$ (upper line), $M_{F}$ (middle line) and $N_{C}$ (lower line). The topological recovery is measured as (A) the Normalised Robinson-Foulds metric and (B) the Normalised Triplets metric calculated using the Bhattacharyya Coefficient. The Bhattacharyya Coefficient values are indicated using a colour gradient ranging from low probability of overlap in blue, to high probability of overlap in orange. Blue regions denote a poor overlap in Normalised metric values between the different parameter combinations (i.e. the parameters have a strong effect on the metric and thus the topological recovery). Conversely, orange regions denote a high overlap in Normalised metric values between the different parameter combinations (i.e. the parameters have a weak effect on the metric and thus the topological recovery).\\

@@

\noindent
\textbf{Table 1:} Bhattacharyya Coefficients of the pairwise method comparisons.
Each line summarises the probabilities of overlap between the distributions of the ``best'' tree versus trees from each inference method (Maximum Likelihood; Bayesian consensus; Maximum Likelihood Bootstraps and Bayesian posterior trees) pooled across all combinations of missing data parameter values, using the Normalised Robinson-Foulds (RF) and Triplets (Tr) metrics. 
Values highlighted in bold are the extreme values of high or low probability of overlap between two methods. If two methods have a high probability of overlap, they have a similar ability to recover the ``correct'' tree topology.
Values $>$ 0.95 denote significantly similar distributions and values $<$ 0.05 denote significantly different distributions.\\
